---
title: "Reproducible Research Homework Assignment 1"
author: "Monobot"
date: "September 17, 2015"
output:
  html_document:
    fig_caption: yes
    keep_md: yes
    self_contained: no
---

please note: a majority of the notes describing each section is either summarized, or copied verbatum from the coursera reproducible research homework assignment. 

###Load and preprocess data

```{r, echo=TRUE}

#import libraries
library(plyr)
library(dplyr)
library(chron)
library(lattice)
library(ggplot2)

#load
rawData <- read.csv('activity.csv')

#remove NAs
cleanData <- na.omit(rawData)

#I may still need some more processing and transforming steps here

```


###What is mean total number of steps taken per day?

Ignoring the missing values in the dataset:

    1. Calculate the total number of steps taken per day

    2. Make a histogram of the total number of steps taken each day

    3. Calculate and report the mean and median of the total number of steps taken per day


```{r, echo=TRUE}

# 1) Calculate total number of steps taken per day

stepsPerDay <-ddply(cleanData, .(date), summarize, sum=sum(steps))

# 2) histogram total number of steps taken per day
par(mfrow=c(1,1))

hist(stepsPerDay$sum, main="Number of Steps/Day", xlab="steps", ylab="frequency", length(stepsPerDay$date))

# 3) Calculate and report the mean and median of the total number of steps taken per day

#mean daily steps
meanDailySteps <-ddply(cleanData, .(date), summarize, mean=mean(steps))

#median daily steps
medianDailySteps <-ddply(cleanData, .(date), summarize, median=quantile(steps,1))

#merge mean and median to single data frame
meanMedianDailySteps <- merge(meanDailySteps,medianDailySteps,by="date")

meanMedianDailySteps

```

###What is the average daily activity pattern?
1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r, echo=TRUE}
#create new data frame with mean values accross days for each time interval
meanStepsPerInterval <- ddply(cleanData, .(interval), summarize, mean=mean(steps))

par(mfrow=c(1,1))

#make time series plot 
plot(meanStepsPerInterval$interval, meanStepsPerInterval$mean, type="l", main="Average Steps Per Interval", ylab="# of steps", xlab="time (in 5 min intervals)")

#find time interval with max average number of steps 
max(meanStepsPerInterval$mean) 
#206.1698 or exactly: 206.1698113

meanStepsPerInterval[which(meanStepsPerInterval$mean == max(meanStepsPerInterval$mean)), ] 
#answer is 835, mean: 206.1698 

```

###Imputing missing values

    1. Calculate and report the total number of missing values in the dataset

    2. Using the strategy described below to fill in missing data, create a new dataset that is equal to the original dataset but with the missing data filled in.
    
    Strategy: 
    I chose to simply add the average number of steps accross all times and days to each NA value.
    Although this is probably the worst strategy, it demonstrates the concept of imputting data.
    
    The impact of adding the artificial values is that we some days that were previously left out due to having NA's are now included; and, for both mean and median, the values have flucutated higher and lower.

   In the histogram, and the new data presented you can see the effect of imputing the artificial data.

```{r, echo=TRUE}
#1 find number of rows containing NAs in raw data
sum(is.na(rawData$steps)) 
#2304

#2 & 3
#first copy rawData to new data frame to manipulate
modData <- rawData

#imputting data steps:
#get vector of NA's
#where value = NA, enter "mean" from meanStepsPerInterval at that interval 
#fill in with the average number of steps
modData$steps[which(is.na(rawData$steps))] <- mean(modData$steps, na.rm = TRUE)

stepsPerDayMod <-ddply(modData, .(date), summarize, sum=sum(steps))

par(mfrow=c(1,2))

#new graph with artificial data added
hist(stepsPerDayMod$sum, main="Number of Steps/Day \n(Artificial NA values added)", xlab="steps", ylab="frequency", ylim=c(0,12), length(stepsPerDayMod$date))

#orginal for comparison
hist(stepsPerDay$sum, main="Number of Steps/Day \n(Original Plot)", xlab="steps", ylab="frequency", ylim=c(0,12), length(stepsPerDay$date))

#reset graphics to 1
par(mfrow=c(1,1))

#4
#mean daily steps
meanDailyStepsMod <-ddply(modData, .(date), summarize, mean=mean(steps))

#median daily steps
medianDailyStepsMod <-ddply(modData, .(date), summarize, median=quantile(steps,1))

#merge mean and median to single data frame
meanMedianDailyStepsMod <- merge(meanDailyStepsMod,medianDailyStepsMod,by="date")

meanMedianDailyStepsMod

#unmodified:
#meanMedianDailySteps

```


###Are there differences in activity patterns between weekdays and weekends?

    Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

    Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.
    
```{r, echo=TRUE}
#create new dataset
modData2 <- modData

#set all of new column = FALSE
modData2$weekend <- FALSE

#set only weekends to "TRUE"
modData2$weekend[which(is.weekend(modData$date))] <- TRUE

#create 2 new dataframes
weekEnd <- modData2[modData2$weekend == TRUE, ]

weekDay <- modData2[modData2$weekend == FALSE, ]

#find mean of each day/weekday or weekend
weekEndMeanSteps <- ddply(weekEnd, .(interval), summarize, mean=mean(steps))
weekDayMeanSteps <- ddply(weekDay, .(interval), summarize, mean=mean(steps))

#add factor variable
weekEndMeanSteps$day <- "weekend"
weekDayMeanSteps$day <- "weekday"

#merge to 1 new dataframe
mergedDF <- rbind(weekEndMeanSteps, weekDayMeanSteps)

#plot graph
p <- ggplot(mergedDF,aes(x=interval, y=mean)) 
p <- p + geom_line() + facet_grid(day ~ .) + theme_bw()
p <- p + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

p
```